\textbf{Transparency to hybrid open access through publisher-provided metadata: An article-level study of Elsevier}

### 2021-05-02 analysis/manuscript.tex
starting revision

The problem with this lack of transparency becomes more apparent in the
case of hybrid OA. In this OA business model, individual articles can be
made openly available for a fee, also known as article-processing charge
(APC), while the journal as a whole remains subscription access. Hence,
hybrid OA was originally introduced as a low-risk transitional model
that allows journals to gradually convert to full OA and reduce
subscription costs as the OA uptake increases (Prosser, 2003). Many
large subscription publishers have since incorporated this OA model, and
while the uptake among authors increased only slowly during the first
decade (Björk, 2012), it has gradually gained popularity in recent
years. This development was driven by the implementation of OA policies
by research institutions and funders, the allocation of OA funds, and
formal agreements between publishers and institutions that allow
affiliated authors to publish OA free-of-charge (Laakso \& Björk, 2016).
However, there have been concerns that the hybrid model allows for
double-dipping---receiving two payments for one article, the APC and
subscription fees (Prosser, 2015; Shieber, 2009). While publishers
assure they combine subscription fees and APCs to adjust pricing
(Mittermaier, 2015), without transparency around the uptake of hybrid OA
and revenue streams, such claims are impossible to evaluate.

@@

The ability to identify and discover hybrid OA articles plays a
fundamental role in monitoring hybrid OA, yet time and again, previous
studies have underscored the difficulties involved in this
(cf.~(Akbaritabar \& Stahlschmidt, 2019; Björk, 2012; Emery, 2013;
Laakso \& Björk, 2016; Leeuwen et al., 2018). Insights into the volume
of hybrid OA and the various levels and sources of APC funding could
create a better understanding of the costs involved in scholarly
publishing and, thus, strengthen institutions' bargaining power in
negotiations with publishers, and promote greater equity in scholarly
publishing. In the following section, we examine how previous research
has tackled the challenge of identifying hybrid OA content and what
these studies taught us about the uptake and funding sources of hybrid
OA.

@@

Since its launch, Unpaywall has become widely used as an OA evidence
source in bibliometric research and rankings(Huang et al., 2020;
Robinson-Garcia et al., 2020). However, results cannot easily be
compared with each other over time because of Unpaywall's ongoing
changes due to non-standardized data to discover hybrid OA. These
changes mainly reflect the growing albeit incomprehensive license
information in Crossref metadata. To overcome these shortcomings, the
developers extended previous web-scraping approaches to better
distinguish between immediate and delayed OA based on licenses at the
date of observation Piwowar et al. (2019). Relatedly, in October 2020,
Elsevier's license for delayed OA content was added to Unpaywall's list
of license patterns (Unpaywall, n.d.-c). At the journal-level, a more
fine-grained identification of full OA journals using OA article
information was developed to better distinguish them from subscription
journals including hybrid journals (Unpaywall, n.d.-b). With its servie
unsub, Unpaywall recently launched a paid service, where libraries can
compare hybrid OA uptake rates with internal financial and usage data to
assess the value of journal subscriptions (Chawla, 2020).

@@

Some countries like Sweden and the United Kingdom (UK) have negotiated
national publishing agreements that allow authors to publish hybrid OA
at a discounted fee or for free. These agreements can serve as the basis
for studies on the uptake of hybrid OA or cost-benefit analyses
(cf.~Olsson (2018); Lawson (2017)) but are often limited in scope as
they tend to focus on a single country and rely on data that seldom is
publicly available. Hence, while being valuable for specific use-cases,
such studies can neither provide a comprehensive understanding of how
the scholarly publishing system functions beyond national boundaries nor
facilitate collective sharing of financial information about hybrid OA.

@@

Next, we used an openly available Crossref database snapshot (Crossref,
2020), which contains all Crossref records registered until March 2020,
to calculate the journals' combined article volume for the five-year
period from 2015-2019. We only included articles published in regular
issues aside from supplements containing conference contributions like
meeting abstracts, indicated by non-numeric pagination. We excluded
non-scholarly journal content, such as the table of contents and lists
of reviewers, following Unpaywall's paratext recognition approach
(Unpaywall, n.d.-a), which we expanded to include patterns indicating
corrections. Furthermore, we categorized the articles by subject
according to the All Science Journal Classification code (ASJC) and
added citation indicators from the Scopus Source Title list (version
June 2020) based on the journal they were published in.

@@

Moreover, we obtained APC invoicing metadata from the XML (see Table 1),
which enabled us to identify the invoice channels and recipients of
hybrid OA APCs. Based on the \texttt{openaccessSponsorType} node in the
XML files, we distinguish between four invoice channels, including
invoices billed to the corresponding author, as part of publishing
agreements with funding bodies (hereinafter ``agreements''; cf.~Elsevier
(n.d.-a)), fee waivers (e.g., in ``cases of genuine need'' or due to
society or university sponsorships, cf.~ref), and other types not
specified by Elsevier. Finally, when hybrid APCs were invoiced as part
of OA agreements, we identified invoice recipients through the
\texttt{openaccessSponsorName} node.

We manually classified invoice recipients based on their institutional
sectors, countries, and primary research areas. Following the OECD's
Frascati Manual (OECD, 2015, p. 91), we coded for four sectors, business
enterprise, government, higher education, and private non-profit. The
government sector consists of federal, state, and municipal governments,
plus non-profit institutions controlled by government units (OECD, 2015,
p. 100). The higher education sector comprises institutions that perform
research and development and provide formal tertiary education services,
such as universities or research institutes (OECD, 2015, p. 101). Legal
or social entities that produce goods and services without generating
profit, such as learned societies or charities, constitute the private
non-profit sector (OECD, 2015, p. 288). Due to the low article volume,
we combined invoice recipients Elsevier listed as ``authors'' and
``third-party sponsor'', and those we had coded as business enterprises
(OECD, 2015, p. 97), into ``Others''. Moreover, we categorized invoice
recipients according to the country or countries representing their
scope of funding and based on the following primary research areas
health sciences, life sciences, physical sciences and mathematics,
social sciences and humanities, unknown, and broad (i.e., multiple
research areas).

@@

Our overall investigation of invoice channels and recipients of hybrid
OA APCs found that the majority of Elsevier hybrid articles were
invoiced to the author (n=41,725; 58.2\%). However, it is important to
note that although Elsevier's metadata classifies the sponsor type as
``author'', this does not necessarily mean that authors paid for APCs
out of their own pocket but rather just that the APC was not invoiced
through an agreement. Hence, it is possible that APCs invoiced to the
author were covered by institutional funds or as part of research
grants. For instance, a recent Springer Nature survey found that hybrid
OA was predominantly supported through institutional and funder sources
(71\%), followed by OA agreements (34\%), while only 6\% were paid from
personal funds or savings (Monaghan et al., 2020). We also observed
notable differences in licensing among invoice types. The majority of
hybrid OA articles, regardless of the invoice type, were licensed under
the more restrictive CC BY-NC-ND license. There is a lack of dedicated
studies on licence selection, but indications so far suggest that when
given a choice authors tend to select more restrictive variants for
their published works (Fraser et al., 2020; Noorden, 2013; Rowley et
al., 2017). On the other hand, when hybrid articles were invoiced
through agreements, most articles were licensed under the more liberal
CC BY license. As several funding bodies mandate CC BY licenses,
including the UK research funders that account for 62\% of our
``agreement'' subsample, this result is perhaps not surprising but
suggests the effectiveness of such agreements.

Based on our findings, around a third of the articles were invoiced
directly through OA agreements (e.g., research funders, national library
consortia). Strikingly, sponsors based in the UK accounted for more than
half of these , and a quarter was charged to sponsors from other
European countries. This finding reiterates reports from previous
studies that the UK's OA profile differs markedly from other countries
(cf.~Jubb et al. (2015); Jubb et al. (2017)) and points to the effects
of science policy. To increase OA to research publications, the UK
implemented centralized APC funding and embraced hybrid OA as a
transition model. Within five years of the publication of the Finch
Report in 2012, the UK recorded an 18\% increase in immediate OA,
coupled with a rise in hybrid OA from 2.7\% in 2012 to 15.4\% of all
articles in 2016 (Jubb et al., 2017), which can be attributed to the
availability of APC funding from Research Councils UK (RCUK). Indeed,
RCUK block grants have been the largest single source of OA funds in the
UK---the Wellcome Trust, and more recently transformative agreements
(Jubb et al., 2017; Tickel, 2018). However, since then publishing
expenditures of UK universities have been rising rapidly (ref), so that
several universities stopped supporting hybrid OA in late 2018-2019
(Castro, 2019; University of Birmingham, n.d.; University of Oxford,
2019; Walker, 2019). This development might explain the slight decrease
in hybrid articles invoiced to the UK our data showed around that time
(see Figure \ref{fig:invoice_sponsor_country}).

@@

\hypertarget{refs}{}
\begin{cslreferences}
\leavevmode\hypertarget{ref-Akbaritabar_2019}{}%
Akbaritabar, A., \& Stahlschmidt, S. (2019). \emph{Applying crossref and
unpaywall information to identify gold, hidden gold, hybrid and delayed
open access publications in the KB publication corpus}. SocArXiv.
\url{https://doi.org/10.31235/osf.io/sdzft}

\leavevmode\hypertarget{ref-antelmann_2017}{}%
Antelman, K. (2017). Leveraging the growth of open access in library
collection decision making. \emph{At the helm: leading transformation},
411--422.

@@

\leavevmode\hypertarget{ref-Bj_rk_2010}{}%
Björk, B.-C., Welling, P., Laakso, M., Majlender, P., Hedlund, T., \&
Guðnason, G. (2010). Open access to the scientific journal literature:
Situation 2009. \emph{PLoS ONE}, \emph{5}(6), e11273.
\url{https://doi.org/10.1371/journal.pone.0011273}

@@

\leavevmode\hypertarget{ref-crminer}{}%
Chamberlain, S. (2020). \emph{crminer: Fetch scholary full text from
CrossRef}. \url{https://CRAN.R-project.org/package=crminer}

\leavevmode\hypertarget{ref-rcrossref}{}%
Chamberlain, S., Zhu, H., Jahn, N., Boettiger, C., \& Ram, K. (2020).
\emph{rcrossref: Client for various CrossRef APIs}.
\url{https://CRAN.R-project.org/package=rcrossref}

@@

\leavevmode\hypertarget{ref-Els_Agreements}{}%
Elsevier. (n.d.-a). \emph{Agreements}.
\url{https://web.archive.org/web/20210127152729/https://www.elsevier.com/open-access/agreements}.

\leavevmode\hypertarget{ref-Els_Archive}{}%
Elsevier. (n.d.-b). \emph{Open archive}.
\url{http://web.archive.org/web/20210127201740/https://www.elsevier.com/open-access/open-archive}.

\leavevmode\hypertarget{ref-Els_videos}{}%
Elsevier. (n.d.-c). \emph{Publishing Journey videos: How do I complete
the Rights and Access form?}
\url{https://web.archive.org/web/20210127154144/https://service.elsevier.com/app/answers/detail/a_id/29789/supporthub/publishing/track/APN2ZgoIDv8a~RNiGvwa~yKgpv0qOS75Mv9e~zj~PP_X/}.

\leavevmode\hypertarget{ref-lund}{}%
Elsevier. (n.d.-d). \emph{Publishing options: BIBSAM institute
associated authors}.
\url{https://web.archive.org/web/20210127153520/https://www.ub.lu.se/en/sites/ub.lu.se.en/files/publicering_under_elsevier_avtalet.pdf}.

\leavevmode\hypertarget{ref-vsnu}{}%
Elsevier. (n.d.-e). \emph{Publishing options for: Dutch universities \&
institutes associated authors}.
\url{https://web.archive.org/web/20190530214833/https://www.openaccess.nl/sites/www.openaccess.nl/files/documenten/elsevier_-_vsnu_workflow_updated_18july_v2.pdf}.

\leavevmode\hypertarget{ref-Emery_2013}{}%
Emery, J. (2013). Mining for gold: Identifying the librarians' toolkit
for managing hybrid open access. \emph{Insights: The UKSG Journal},
\emph{26}(2), 115--119. \url{https://doi.org/10.1629/2048-7754.65}

@@

\leavevmode\hypertarget{ref-Fraser_2020}{}%
Fraser, N., Brierley, L., Dey, G., Polka, J. K., Pálfy, M., Nanni, F.,
\& Coates, J. A. (2020). \emph{Preprinting the COVID-19 pandemic}.
bioRxiv. \url{https://doi.org/10.1101/2020.05.22.111294}

@@

\leavevmode\hypertarget{ref-Geschuhn_2017}{}%
Geschuhn, K., \& Stone, G. (2017). It's the workflows, stupid! What is
required to make ``offsetting'' work for the open access transition.
\emph{Insights the UKSG Journal}, \emph{30}(3), 103--114.
\url{https://doi.org/10.1629/uksg.391}

@@

\leavevmode\hypertarget{ref-Huang_2020}{}%
Huang, C.-K. (Karl), Neylon, C., Hosking, R., Montgomery, L., Wilson, K.
S., Ozaygen, A., \& Brookes-Kenworthy, C. (2020). Evaluating the impact
of open access policies on research institutions. \emph{eLife},
\emph{9}. \url{https://doi.org/10.7554/elife.57067}

\leavevmode\hypertarget{ref-Jahn_2016}{}%
Jahn, N., \& Tullney, M. (2016). A study of institutional spending on
open access publication fees in germany. \emph{PeerJ}, \emph{4}, e2323.
\url{https://doi.org/10.7717/peerj.2323}

\leavevmode\hypertarget{ref-Jubb_2015}{}%
Jubb, M., Goldstein, S., Amin, M., Plume, A., Oeben, S., Aisati, M.,
Pinfield, S., Bath, P., Salter, J., Johnson, R., \& Fosci, M. (2015).
\emph{Monitoring the transition to open access: A report for the
universities uk open access co-ordination group}.
\url{https://web.archive.org/web/20190523005240/https://www.acu.ac.uk/research-information-network/monitoring-transition-to-open-access}

@@

\leavevmode\hypertarget{ref-Kirkman_2018}{}%
Kirkman, N. S. (2018). \emph{A study of open access publishing by nhmrc
grant recipients} {[}Curtin University{]}.
\url{http://hdl.handle.net/20.500.11937/77026}

\leavevmode\hypertarget{ref-Kramer_2018}{}%
Kramer, B., \& Bosman, J. (2018). \emph{Towards a plan S gap analysis:
open access potential across disciplines using Web of Science and DOAJ}
{[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.1979937}

@@

\leavevmode\hypertarget{ref-Larivi_re_2015}{}%
Larivière, V., Haustein, S., \& Mongeon, P. (2015). The oligopoly of
academic publishers in the digital era. \emph{PLOS ONE}, \emph{10}(6),
e0127502. \url{https://doi.org/10.1371/journal.pone.0127502}

@@

@@

\leavevmode\hypertarget{ref-Lawson_2015}{}%
Lawson, S. (2015). "Total cost of ownership" of scholarly communication:
Managing subscription and APC payments together. \emph{Learned
Publishing}, \emph{28}(1), 9--13. \url{https://doi.org/10.1087/20150103}

@@

\leavevmode\hypertarget{ref-Marques_2019}{}%
Marques, M., Woutersen-Windhouwer, S., \& Tuuliniemi, A. (2019).
Monitoring agreements with open access elements: Why article-level
metadata are important. \emph{Insights the UKSG Journal}, \emph{32}.
\url{https://doi.org/10.1629/uksg.489}

@@

\leavevmode\hypertarget{ref-Matthias_2020}{}%
Matthias, L. (2020). \emph{Publisher oa portfolios 2.0} (Version 2.0)
{[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.3841568}

@@

\leavevmode\hypertarget{ref-McNutt_2019}{}%
McNutt, M. (2019). Opinion: Plan S falls short for society
publishers--and for the researchers they serve. \emph{Proceedings of the
National Academy of Sciences}, \emph{116}(7), 2400--2403.
\url{https://doi.org/10.1073/pnas.1900359116}

\leavevmode\hypertarget{ref-Mittermaier_2015}{}%
Mittermaier, B. (2015). Double dipping in hybrid open access -- chimera
or reality? \emph{ScienceOpen Research}.
\url{https://doi.org/10.14293/s2199-1006.1.sor-socsci.aowntu.v1}

@@

\leavevmode\hypertarget{ref-Frascati}{}%
OECD. (2015). \emph{Frascati manual 2015: Guidelines for collecting and
reporting data on research and experimental development}. OECD
Publishing. \url{https://doi.org/10.1787/24132764}

@@

\leavevmode\hypertarget{ref-Pieper_2018}{}%
Pieper, D., \& Broschinski, C. (2018). OpenAPC: A contribution to a
transparent and reproducible monitoring of fee-based open access
publishing across institutions and nations. \emph{Insights the UKSG
Journal}, \emph{31}. \url{https://doi.org/10.1629/uksg.439}

@@

\leavevmode\hypertarget{ref-Piwowar_2018}{}%
Piwowar, H., Priem, J., Larivière, V., Alperin, J. P., Matthias, L.,
Norlander, B., Farley, A., West, J., \& Haustein, S. (2018). The state
of OA: A large-scale analysis of the prevalence and impact of open
access articles. \emph{PeerJ}, \emph{6}, e4375.
\url{https://doi.org/10.7717/peerj.4375}

\leavevmode\hypertarget{ref-Piwowar_2019}{}%
Piwowar, H., Priem, J., \& Orr, R. (2019). \emph{The future of OA: A
large-scale analysis projecting open access publication and readership}.
bioRxiv. \url{https://doi.org/10.1101/795310}

@@

\leavevmode\hypertarget{ref-Prosser_2015}{}%
Prosser, D. (2015). \emph{The costs of double dipping}.
\url{https://web.archive.org/web/20210128164930/https://www.rluk.ac.uk/the-costs-of-double-dipping/}.

@@

\leavevmode\hypertarget{ref-Robinson_Garcia_2020}{}%
Robinson-Garcia, N., Costas, R., \& Leeuwen, T. N. van. (2020). Open
access uptake by universities worldwide. \emph{PeerJ}, \emph{8}, e9410.
\url{https://doi.org/10.7717/peerj.9410}

>>>>

The problem with this lack of transparency becomes more apparent in the
case of hybrid OA. In this OA business model, individual articles can be
made openly available for a fee, also known as article-processing charge
(APC), while the journal as a whole remains subscription access. Hence,
hybrid OA was originally introduced as a low-risk transitional model
that allows journals to gradually convert to full OA and reduce
subscription costs as the OA uptake increases (D. C. Prosser, 2003).
Many large subscription publishers have since incorporated this OA
model, and while the uptake among authors increased only slowly during
the first decade (Björk, 2012), it has gradually gained popularity in
recent years. This development was driven by the implementation of OA
policies by research institutions and funders, the allocation of OA
funds, and formal agreements between publishers and institutions that
allow affiliated authors to publish OA free-of-charge (Laakso \& Björk,
2016). However, there have been concerns that the hybrid model allows
for double-dipping---receiving two payments for one article, the APC and
subscription fees (D. Prosser, 2015; Shieber, 2009). While publishers
assure they combine subscription fees and APCs to adjust pricing
(Mittermaier, 2015), without transparency around the uptake of hybrid OA
and revenue streams, such claims are impossible to evaluate.

@@

The ability to identify and discover hybrid OA articles plays a
fundamental role in monitoring hybrid OA, yet time and again, previous
studies have underscored the difficulties involved in this (cf.
(Akbaritabar \& Stahlschmidt, 2019; Björk, 2012; Emery, 2013; Laakso \&
Björk, 2016; Leeuwen et al., 2018). Insights into the volume of hybrid
OA and the various levels and sources of APC funding could create a
better understanding of the costs involved in scholarly publishing and,
thus, strengthen institutions' bargaining power in negotiations with
publishers, and promote greater equity in scholarly publishing. In the
following section, we examine how previous research has tackled the
challenge of identifying hybrid OA content and what these studies taught
us about the uptake and funding sources of hybrid OA.

@@

Since its launch, Unpaywall has become widely used as an OA evidence
source in bibliometric research and rankings(Huang et al., 2020;
Robinson-Garcia et al., 2020). However, results cannot easily be
compared with each other over time because of Unpaywall's ongoing
changes due to non-standardized data to discover hybrid OA. These
changes mainly reflect the growing albeit incomprehensive license
information in Crossref metadata. To overcome these shortcomings, the
developers extended previous web-scraping approaches to better
distinguish between immediate and delayed OA based on licenses at the
date of observation Piwowar et al. (2019). Relatedly, in October 2020,
Elsevier's license for delayed OA content was added to Unpaywall's list
of license patterns (Unpaywall, n.d.-c). At the journal-level, a more
fine-grained identification of full OA journals using OA article
information was developed to better distinguish them from subscription
journals including hybrid journals (Unpaywall, n.d.-a). With its servie
unsub, Unpaywall recently launched a paid service, where libraries can
compare hybrid OA uptake rates with internal financial and usage data to
assess the value of journal subscriptions (Chawla, 2020).

@@

Some countries like Sweden and the United Kingdom (UK) have negotiated
national publishing agreements that allow authors to publish hybrid OA
at a discounted fee or for free. These agreements can serve as the basis
for studies on the uptake of hybrid OA or cost-benefit analyses (cf.
Olsson (2018); Lawson (2017)) but are often limited in scope as they
tend to focus on a single country and rely on data that seldom is
publicly available. Hence, while being valuable for specific use-cases,
such studies can neither provide a comprehensive understanding of how
the scholarly publishing system functions beyond national boundaries nor
facilitate collective sharing of financial information about hybrid OA.

@@

Next, we used an openly available Crossref database snapshot (Crossref,
2020), which contains all Crossref records registered until March 2020,
to calculate the journals' combined article volume for the five-year
period from 2015-2019. We only included articles published in regular
issues aside from supplements containing conference contributions like
meeting abstracts, indicated by non-numeric pagination. We excluded
non-scholarly journal content, such as the table of contents and lists
of reviewers, following Unpaywall's paratext recognition approach
(Unpaywall, n.d.-b), which we expanded to include patterns indicating
corrections. Furthermore, we categorized the articles by subject
according to the All Science Journal Classification code (ASJC) and
added citation indicators from the Scopus Source Title list (version
June 2020) based on the journal they were published in.

@@

Moreover, we obtained APC invoicing metadata from the XML (see Table 1),
which enabled us to identify the invoice channels and recipients of
hybrid OA APCs. Based on the \texttt{openaccessSponsorType} node in the
XML files, we distinguish between four invoice channels, including
invoices billed to the corresponding author, as part of publishing
agreements with funding bodies (hereinafter ``agreements''; cf. Elsevier
(n.d.-a)), fee waivers (e.g., in ``cases of genuine need'' or due to
society or university sponsorships, cf.~ref), and other types not
specified by Elsevier. Finally, when hybrid APCs were invoiced as part
of OA agreements, we identified invoice recipients through the
\texttt{openaccessSponsorName} node.

We manually classified invoice recipients based on their institutional
sectors, countries, and primary research areas. Following the OECD's
Frascati Manual (OECD, 2015, p. 91), we coded for four sectors, business
enterprise, government, higher education, and private non-profit. The
government sector consists of federal, state, and municipal governments,
plus non-profit institutions controlled by government units (OECD, 2015,
p. 100). The higher education sector comprises institutions that perform
research and development and provide formal tertiary education services,
such as universities or research institutes (OECD, 2015, p. 101). Legal
or social entities that produce goods and services without generating
profit, such as learned societies or charities, constitute the private
non-profit sector (OECD, 2015, p. 288). Due to the low article volume,
we combined invoice recipients Elsevier listed as ``authors'' and
``third-party sponsor,'' and those we had coded as business enterprises
(OECD, 2015, p. 97), into ``Others.'' Moreover, we categorized invoice
recipients according to the country or countries representing their
scope of funding and based on the following primary research areas
health sciences, life sciences, physical sciences and mathematics,
social sciences and humanities, unknown, and broad (i.e., multiple
research areas).

@@

Our overall investigation of invoice channels and recipients of hybrid
OA APCs found that the majority of Elsevier hybrid articles were
invoiced to the author (n=41,725; 58.2\%). However, it is important to
note that although Elsevier's metadata classifies the sponsor type as
``author,'' this does not necessarily mean that authors paid for APCs
out of their own pocket but rather just that the APC was not invoiced
through an agreement. Hence, it is possible that APCs invoiced to the
author were covered by institutional funds or as part of research
grants. For instance, a recent Springer Nature survey found that hybrid
OA was predominantly supported through institutional and funder sources
(71\%), followed by OA agreements (34\%), while only 6\% were paid from
personal funds or savings (Monaghan et al., 2020). We also observed
notable differences in licensing among invoice types. The majority of
hybrid OA articles, regardless of the invoice type, were licensed under
the more restrictive CC BY-NC-ND license. There is a lack of dedicated
studies on licence selection, but indications so far suggest that when
given a choice authors tend to select more restrictive variants for
their published works (Fraser et al., 2020; Noorden, 2013; Rowley et
al., 2017). On the other hand, when hybrid articles were invoiced
through agreements, most articles were licensed under the more liberal
CC BY license. As several funding bodies mandate CC BY licenses,
including the UK research funders that account for 62\% of our
``agreement'' subsample, this result is perhaps not surprising but
suggests the effectiveness of such agreements.

Based on our findings, around a third of the articles were invoiced
directly through OA agreements (e.g., research funders, national library
consortia). Strikingly, sponsors based in the UK accounted for more than
half of these , and a quarter was charged to sponsors from other
European countries. This finding reiterates reports from previous
studies that the UK's OA profile differs markedly from other countries
(cf. Jubb et al. (2015); Jubb et al. (2017)) and points to the effects
of science policy. To increase OA to research publications, the UK
implemented centralized APC funding and embraced hybrid OA as a
transition model. Within five years of the publication of the Finch
Report in 2012, the UK recorded an 18\% increase in immediate OA,
coupled with a rise in hybrid OA from 2.7\% in 2012 to 15.4\% of all
articles in 2016 (Jubb et al., 2017), which can be attributed to the
availability of APC funding from Research Councils UK (RCUK). Indeed,
RCUK block grants have been the largest single source of OA funds in the
UK---the Wellcome Trust, and more recently transformative agreements
(Jubb et al., 2017; Tickel, 2018). However, since then publishing
expenditures of UK universities have been rising rapidly (ref), so that
several universities stopped supporting hybrid OA in late 2018-2019
(Castro, 2019; University of Birmingham, n.d.; University of Oxford,
2019; Walker, 2019). This development might explain the slight decrease
in hybrid articles invoiced to the UK our data showed around that time
(see Figure \ref{fig:invoice_sponsor_country}).

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-Akbaritabar_2019}{}%
Akbaritabar, A., \& Stahlschmidt, S. (2019). \emph{Applying crossref and
unpaywall information to identify gold, hidden gold, hybrid and delayed
open access publications in the {KB} publication corpus}. SocArXiv.
\url{https://doi.org/10.31235/osf.io/sdzft}

\leavevmode\hypertarget{ref-antelmann_2017}{}%
Antelman, K. (2017). Leveraging the growth of open access in library
collection decision making. \emph{{At the helm: leading
transformation}}, 411--422.

@@

\leavevmode\hypertarget{ref-Bj_rk_2010}{}%
Björk, B.-C., Welling, P., Laakso, M., Majlender, P., Hedlund, T., \&
Guðnason, G. (2010). Open access to the scientific journal literature:
Situation 2009. \emph{{PLoS} {ONE}}, \emph{5}(6), e11273.
\url{https://doi.org/10.1371/journal.pone.0011273}

@@

\leavevmode\hypertarget{ref-crminer}{}%
Chamberlain, S. (2020). \emph{{crminer}: Fetch scholary full text from
{CrossRef}}. \url{https://CRAN.R-project.org/package=crminer}

\leavevmode\hypertarget{ref-rcrossref}{}%
Chamberlain, S., Zhu, H., Jahn, N., Boettiger, C., \& Ram, K. (2020).
\emph{{rcrossref}: Client for various {CrossRef} {APIs}}.
\url{https://CRAN.R-project.org/package=rcrossref}

@@

\leavevmode\hypertarget{ref-Els_Agreements}{}%
Elsevier. (n.d.-a). \emph{{Agreements}}.
\url{https://web.archive.org/web/20210127152729/https://www.elsevier.com/open-access/agreements}.

\leavevmode\hypertarget{ref-Els_Archive}{}%
Elsevier. (n.d.-b). \emph{{Open archive}}.
\url{http://web.archive.org/web/20210127201740/https://www.elsevier.com/open-access/open-archive}.

\leavevmode\hypertarget{ref-Els_videos}{}%
Elsevier. (n.d.-c). \emph{{Publishing Journey videos: How do I complete
the Rights and Access form?}}
\url{https://web.archive.org/web/20210127154144/https://service.elsevier.com/app/answers/detail/a_id/29789/supporthub/publishing/track/APN2ZgoIDv8a~RNiGvwa~yKgpv0qOS75Mv9e~zj~PP_X/}.

\leavevmode\hypertarget{ref-lund}{}%
Elsevier. (n.d.-d). \emph{{Publishing options: BIBSAM institute
associated authors}}.
\url{https://web.archive.org/web/20210127153520/https://www.ub.lu.se/en/sites/ub.lu.se.en/files/publicering_under_elsevier_avtalet.pdf}.

\leavevmode\hypertarget{ref-vsnu}{}%
Elsevier. (n.d.-e). \emph{{Publishing options for: Dutch universities \&
institutes associated authors}}.
\url{https://web.archive.org/web/20190530214833/https://www.openaccess.nl/sites/www.openaccess.nl/files/documenten/elsevier_-_vsnu_workflow_updated_18july_v2.pdf}.

\leavevmode\hypertarget{ref-Emery_2013}{}%
Emery, J. (2013). Mining for gold: Identifying the librarians' toolkit
for managing hybrid open access. \emph{Insights: The {UKSG} Journal},
\emph{26}(2), 115--119. \url{https://doi.org/10.1629/2048-7754.65}

@@

\leavevmode\hypertarget{ref-Fraser_2020}{}%
Fraser, N., Brierley, L., Dey, G., Polka, J. K., Pálfy, M., Nanni, F.,
\& Coates, J. A. (2020). \emph{Preprinting the {COVID}-19 pandemic}.
{bioRxiv}. \url{https://doi.org/10.1101/2020.05.22.111294}

@@

\leavevmode\hypertarget{ref-Geschuhn_2017}{}%
Geschuhn, K., \& Stone, G. (2017). It's the workflows, stupid! What is
required to make {`offsetting'} work for the open access transition.
\emph{Insights the {UKSG} Journal}, \emph{30}(3), 103--114.
\url{https://doi.org/10.1629/uksg.391}

@@

\leavevmode\hypertarget{ref-Huang_2020}{}%
Huang, C.-K. (Karl), Neylon, C., Hosking, R., Montgomery, L., Wilson, K.
S., Ozaygen, A., \& Brookes-Kenworthy, C. (2020). Evaluating the impact
of open access policies on research institutions. \emph{{eLife}},
\emph{9}. \url{https://doi.org/10.7554/elife.57067}

\leavevmode\hypertarget{ref-Jahn_2016}{}%
Jahn, N., \& Tullney, M. (2016). A study of institutional spending on
open access publication fees in germany. \emph{{PeerJ}}, \emph{4},
e2323. \url{https://doi.org/10.7717/peerj.2323}

\leavevmode\hypertarget{ref-Jubb_2015}{}%
Jubb, M., Goldstein, S., Amin, M., Plume, A., Oeben, S., Aisati, M.,
Pinfield, S., Bath, P., Salter, J., Johnson, R., \& Fosci, M. (2015).
\emph{Monitoring the transition to open access: A report for the
universities UK open access co-ordination group}.
\url{https://web.archive.org/web/20190523005240/https://www.acu.ac.uk/research-information-network/monitoring-transition-to-open-access}

@@

\leavevmode\hypertarget{ref-Kirkman_2018}{}%
Kirkman, N. S. (2018). \emph{A study of open access publishing by NHMRC
grant recipients} {[}{Curtin University}{]}.
\url{http://hdl.handle.net/20.500.11937/77026}

\leavevmode\hypertarget{ref-Kramer_2018}{}%
Kramer, B., \& Bosman, J. (2018). \emph{{Towards a plan S gap analysis:
open access potential across disciplines using Web of Science and DOAJ}}
{[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.1979937}

@@

\leavevmode\hypertarget{ref-Larivi_re_2015}{}%
Larivière, V., Haustein, S., \& Mongeon, P. (2015). The oligopoly of
academic publishers in the digital era. \emph{{PLOS} {ONE}},
\emph{10}(6), e0127502.
\url{https://doi.org/10.1371/journal.pone.0127502}

@@

\leavevmode\hypertarget{ref-Lawson_2015}{}%
Lawson, S. (2015). "Total cost of ownership" of scholarly communication:
Managing subscription and {APC} payments together. \emph{Learned
Publishing}, \emph{28}(1), 9--13. \url{https://doi.org/10.1087/20150103}

@@

@@

\leavevmode\hypertarget{ref-Marques_2019}{}%
Marques, M., Woutersen-Windhouwer, S., \& Tuuliniemi, A. (2019).
Monitoring agreements with open access elements: Why article-level
metadata are important. \emph{Insights the {UKSG} Journal}, \emph{32}.
\url{https://doi.org/10.1629/uksg.489}

@@

\leavevmode\hypertarget{ref-Matthias_2020}{}%
Matthias, L. (2020). \emph{Publisher OA portfolios 2.0} (Version 2.0)
{[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.3841568}

@@

\leavevmode\hypertarget{ref-McNutt_2019}{}%
McNutt, M. (2019). Opinion: {Plan S} falls short for society
publishers--and for the researchers they serve. \emph{Proceedings of the
National Academy of Sciences}, \emph{116}(7), 2400--2403.
\url{https://doi.org/10.1073/pnas.1900359116}

\leavevmode\hypertarget{ref-Mittermaier_2015}{}%
Mittermaier, B. (2015). Double dipping in hybrid open access -- chimera
or reality? \emph{{ScienceOpen} Research}.
\url{https://doi.org/10.14293/s2199-1006.1.sor-socsci.aowntu.v1}

@@

\leavevmode\hypertarget{ref-Frascati}{}%
OECD. (2015). \emph{Frascati manual 2015: Guidelines for collecting and
reporting data on research and experimental development}. {OECD
Publishing}. \url{https://doi.org/10.1787/24132764}

@@

\leavevmode\hypertarget{ref-Pieper_2018}{}%
Pieper, D., \& Broschinski, C. (2018). {OpenAPC}: A contribution to a
transparent and reproducible monitoring of fee-based open access
publishing across institutions and nations. \emph{Insights the {UKSG}
Journal}, \emph{31}. \url{https://doi.org/10.1629/uksg.439}

@@

\leavevmode\hypertarget{ref-Piwowar_2018}{}%
Piwowar, H., Priem, J., Larivière, V., Alperin, J. P., Matthias, L.,
Norlander, B., Farley, A., West, J., \& Haustein, S. (2018). The state
of {OA}: A large-scale analysis of the prevalence and impact of open
access articles. \emph{{PeerJ}}, \emph{6}, e4375.
\url{https://doi.org/10.7717/peerj.4375}

\leavevmode\hypertarget{ref-Piwowar_2019}{}%
Piwowar, H., Priem, J., \& Orr, R. (2019). \emph{The future of {OA}: A
large-scale analysis projecting open access publication and readership}.
{bioRxiv}. \url{https://doi.org/10.1101/795310}

@@

\leavevmode\hypertarget{ref-Prosser_2015}{}%
Prosser, D. (2015). \emph{{The costs of double dipping}}.
\url{https://web.archive.org/web/20210128164930/https://www.rluk.ac.uk/the-costs-of-double-dipping/}.

@@

\leavevmode\hypertarget{ref-Robinson_Garcia_2020}{}%
Robinson-Garcia, N., Costas, R., \& Leeuwen, T. N. van. (2020). Open
access uptake by universities worldwide. \emph{{PeerJ}}, \emph{8},
e9410. \url{https://doi.org/10.7717/peerj.9410}
